---
title: "Forecasting US National Accounts: GDP, Domestic Demand, and Imports"
author: "Noah K and Crawford H"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
editor: visual
---

# AI Disclaimer:

AI, specifically Claude, Perplexity, and Gemini were consulted with to troubleshoot errors, form tables for the output values, and help with cleaning up our code. AI also helped us to add additions making our knitted report very pretty.

```{r setup}
#| label: setup
#| echo: true
#| message: false
#| warning: false

rm(list = ls())

library(tidyverse)
library(tsibble)
library(fpp3)
library(fredr)
library(scales)
library(vars)
library(fable)
library(feasts)
library(knitr)
library(kableExtra)
```

# Data Loading and Preparation

## Load Raw CSV Files

```{r load-data}
#| echo: true
#| message: false
#| warning: false

# OECD data: US GDP, Domestic Demand, and Imports of Goods & Services
# Period: 2000 Q1 to 2025 Q4
# Units: national currency, chain-linked volume, base year = 2017
# Unit multiplier = 6 (true values = observation × 10^6)

US_DD  <- read.csv("us_dd.csv")
US_IMP <- read.csv("us_imp.csv")
US_GDP <- read.csv("us_gdp.csv")
```

## Clean and Join into Tsibble

Here, the chunk below selects the relevant columns from each CSV and removes the rest, then left joins the tables together to make a singular reference table. This master table is also cleaned, set to a tsibble format for time series analysis, and adds logged versions of each varaible's values for future analysis.

```{r clean-data}
#| echo: true
#| message: false
#| warning: false

US_DD <- US_DD %>%
  dplyr::select(date = TIME_PERIOD, country = REF_AREA,
                domestic_demand = OBS_VALUE)

US_GDP <- US_GDP %>%
  dplyr::select(date = TIME_PERIOD, country = REF_AREA,
                gdp = OBS_VALUE)

US_IMP <- US_IMP %>%
  dplyr::select(date = TIME_PERIOD, country = REF_AREA,
                imports = OBS_VALUE)

us_clean <- US_DD %>%
  left_join(US_GDP, by = c("date", "country")) %>%
  left_join(US_IMP, by = c("date", "country")) %>%
  mutate(date = yearquarter(date)) %>%
  group_by(date, country) %>%
  summarise(
    domestic_demand = sum(domestic_demand, na.rm = TRUE),
    gdp             = sum(gdp,             na.rm = TRUE),
    imports         = sum(imports,         na.rm = TRUE),
    .groups = "drop") %>% as_tsibble(key = country, index = date)

# Add log-transformed columns — stabilises variance and linearises growth
us_clean <- us_clean %>%
  mutate(
    log_gdp             = log(gdp),
    log_domestic_demand = log(domestic_demand),
    log_imports         = log(imports)
  )

glimpse(us_clean)
```

# Exploratory Analysis

## Raw Level Plots

Basic time plots of the base variables, GDP, domestic demand, and import values for the US are created below.

```{r raw-plots}
#| echo: false
#| fig-height: 8

us_clean %>%
  pivot_longer(cols = c(gdp, domestic_demand, imports),
               names_to = "variable", values_to = "value") %>%
  autoplot(value) +
  facet_wrap(~variable, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels = comma) +
  labs(title    = "US National Accounts: Raw Quarterly Levels (2000–2025)",
       subtitle = "Millions of USD, chain-linked volume (2017 base)",
       x = NULL, y = "Value (Millions)") +
  theme_bw()
```

**Analysis:** All three series exhibit clear upward trends with a visible drops around 2008 for the global financial cricis and another drop in 2020 for the COVID pandemic. Imports show the most volatility of the three. The upward drift and different variances across time indicate the logged version of each variable should be taken.

## Log-Transformed Series

Like the previous chunk, we are creating a time plot of each variable, this time for the logged versions of each variable.

```{r log-plots}
#| echo: false
#| fig-height: 8

us_clean %>%
  pivot_longer(cols = c(log_gdp, log_domestic_demand, log_imports),
               names_to = "variable", values_to = "value") %>%
  autoplot(value) +
  facet_wrap(~variable, scales = "free_y", ncol = 1) +
  labs(title    = "Log-Transformed US National Accounts (2000–2025)",
       subtitle = "Log scale linearises exponential growth",
       x = NULL, y = "Log Value") +
  theme_bw()
```

**Analysis:** Log transforming each variable did make the lines smoother, however the strong upward trend remains. The COVID-19 pandemic and the massive drop in value for each variable, will most likey impact the forecast variables. In more statistical terms, the COVID-19 pandemic created outliers in the data here.

## STL Decomposition

```{r stl}
#| echo: false
#| fig-height: 6

us_clean %>%
  model(STL(log_gdp ~ trend(window = 13) + season(window = "periodic"))) %>%
  components() %>% autoplot() +
  labs(title = "STL Decomposition: Log GDP") + theme_bw()

us_clean %>%
  model(STL(log_domestic_demand ~ trend(window = 13) + season(window = "periodic"))) %>%
  components() %>% autoplot() +
  labs(title = "STL Decomposition: Log Domestic Demand") + theme_bw()

us_clean %>%
  model(STL(log_imports ~ trend(window = 13) + season(window = "periodic"))) %>%
  components() %>% autoplot() +
  labs(title = "STL Decomposition: Log Imports") + theme_bw()
```

**Analysis:** All three series share the same decomposition pattern of a clear positive trend, consistent seasonal patterns, and a remainder that hovers around 0 (the mean) until the COVID-19 pandemic. The seasonal component is highly consistent across time, so additive instead of multiplicative seasonality should be present in our ETS models. When we use R and other tests to pick the best ETS decomposition, we will know for sure what specifications are needed.

# Training and Testing Split

As the chunk title implies, we are setting out training data period, 2000 Q1 to 2023 Q4, and our test data period, 2024 Q1 to 2025 Q4.

```{r train-test}
#| echo: true

# Training: 2000 Q1 – 2023 Q4  =  96 quarters
# Test:     2024 Q1 – 2025 Q4  =   8 quarters
us_train <- us_clean %>% filter(date <= yearquarter("2023 Q4"))
us_test  <- us_clean %>% filter(date >= yearquarter("2024 Q1"))

cat("Training observations:", nrow(us_train), "\n")
cat("Test observations:    ", nrow(us_test),  "\n")
```

# ETS Modelling

ETS (Error, Trend, and Seasonal) models decompose a series into three exponentially smoothed components. Since all series are log-transformed, additive ETS on the log scale here is equivalent to multiplicative ETS on the original scale.

## Fit Multiple ETS Specifications

Five specifications are fit per variable: an automatically selected model, two damped/undamped additive variants, and one full additive seasonal model. Residuals are being checked before using the models for forecasting to confirm the data has white-noise for the remainders.

```{r ets-fit}
#| echo: true
#| warning: false

# GDP 
ets_fit_gdp <- us_train %>%
  model(
    auto_ets = ETS(log_gdp),
    ets_AAN  = ETS(log_gdp ~ error("A") + trend("A")  + season("N")),
    ets_AAdN = ETS(log_gdp ~ error("A") + trend("Ad") + season("N")),
    ets_AAA  = ETS(log_gdp ~ error("A") + trend("A")  + season("A")),
    ets_AAdA = ETS(log_gdp ~ error("A") + trend("Ad") + season("A")))

# Domestic Demand
ets_fit_dd <- us_train %>%
  model(
    auto_ets = ETS(log_domestic_demand),
    ets_AAN  = ETS(log_domestic_demand ~ error("A") + trend("A")  + season("N")), ets_AAdN = ETS(log_domestic_demand ~ error("A") + trend("Ad") + season("N")), ets_AAA  = ETS(log_domestic_demand ~ error("A") + trend("A")  + season("A")), ets_AAdA = ETS(log_domestic_demand ~ error("A") + trend("Ad") + season("A")))

# Imports
ets_fit_imp <- us_train %>%
  model(
    auto_ets = ETS(log_imports),
    ets_AAN  = ETS(log_imports ~ error("A") + trend("A")  + season("N")),
    ets_AAdN = ETS(log_imports ~ error("A") + trend("Ad") + season("N")),
    ets_AAA  = ETS(log_imports ~ error("A") + trend("A")  + season("A")),
    ets_AAdA = ETS(log_imports ~ error("A") + trend("Ad") + season("A")))
```

## ETS Model Comparison: AICc and BIC

```{r ets-ic}
#| echo: false

# For ETS of GDP
# AI assisted code to make clearly visible results (like future chunks!)
glance(ets_fit_gdp) %>%
  dplyr::select(.model, AICc, BIC) %>%
  arrange(AICc) %>%
  knitr::kable(
    caption = "ETS for Logged GDP",
    digits = 2,
    col.names = c("Model", "AICc", "BIC"),
    align = c("l", "r", "r")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::row_spec(1, bold = TRUE, background = "#d4edda")

# For ETS of domestic demand
# AI assisted code to make clearly visible results (like future chunks!)
glance(ets_fit_dd) %>%
  dplyr::select(.model, AICc, BIC) %>%
  arrange(AICc) %>%
  knitr::kable(
    caption = "ETS for Logged Domestic Demand",
    digits = 2,
    col.names = c("Model", "AICc", "BIC"),
    align = c("l", "r", "r")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::row_spec(1, bold = TRUE, background = "#d4edda")

# For ETS of imports
# AI assisted code to make clearly visible results (like future chunks!)
glance(ets_fit_imp) %>%
  dplyr::select(.model, AICc, BIC) %>%
  arrange(AICc) %>%
  knitr::kable(
    caption = "ETS for Logged Imports",
    digits = 2,
    col.names = c("Model", "AICc", "BIC"),
    align = c("l", "r", "r")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::row_spec(1, bold = TRUE, background = "#d4edda")

# To see the best ETS model sepcifications
ets_fit_gdp %>%
  dplyr::select(auto_ets, .model_desc = 2) %>%
  print()

ets_fit_dd %>%
  dplyr::select(auto_ets, .model_desc = 2) %>%
  print()

ets_fit_imp %>%
  dplyr::select(auto_ets, .model_desc = 2) %>%
  print()
```

**Analysis:** Lower AICc indicates a better fitting model, and here R's auto-selected model performed the best with every variable. We were right with the first manually selected model of having an additive trend and no seasonality, but no dampening was present in these ETS models. Every auto-selected model by R has the same specifications of ETS(M,A,N), or multiplicative errors, additive trend, and no seasonality.

## ETS Residual Diagnostics

Residuals are checked on the auto-selected model (best AICc candidate) before forecasting. A well-specified model should show 3 main traits: no autocorrelation in the time plot, ACF values within 95% bounds, and a normally distributed histogram.

```{r ets-residuals}
#| echo: false
#| fig-height: 5

ets_fit_gdp %>% dplyr::select(auto_ets) %>%
  gg_tsresiduals() +
  labs(title = "ETS Residual Diagnostics: Log GDP") + theme_bw()

ets_fit_dd %>% dplyr::select(auto_ets) %>%
  gg_tsresiduals() +
  labs(title = "ETS Residual Diagnostics: Log Domestic Demand") + theme_bw()

ets_fit_imp %>% dplyr::select(auto_ets) %>%
  gg_tsresiduals() +
  labs(title = "ETS Residual Diagnostics: Log Imports") + theme_bw()
```

**Analysis:** Examine each panel. The 2020 COVID shock still appears strongly on the time plot, but otherwise the mean is 0 for every variable in the best ETS models. All gg_tsresiduals plots for each ETS have ACF values within the 95% boundary, and a normally distributed histogram. We can therefor conclude that taking the log of the variables has given us stationary data to work with that's white noise. An Ljung-Box test will also be performed.

## Ljung-Box Test on ETS Residuals

The Ljung-Box Test on our ETS models has a null hypothesis that there is no autocorrelation in our variables. If the p-value is below 0.05 as our threshold, we reject the null hypothesis and say that some autocorrelation is present.

```{r ets-ljungbox}
#| echo: false

bind_rows(
  augment(ets_fit_gdp) %>% filter(.model == "auto_ets") %>%
    features(.innov, ljung_box, lag = 8, dof = 0) %>%
    mutate(variable = "Log GDP"),
  augment(ets_fit_dd) %>% filter(.model == "auto_ets") %>%
    features(.innov, ljung_box, lag = 8, dof = 0) %>%
    mutate(variable = "Log Domestic Demand"),
  augment(ets_fit_imp) %>% filter(.model == "auto_ets") %>%
    features(.innov, ljung_box, lag = 8, dof = 0) %>%
    mutate(variable = "Log Imports")
) %>%
  dplyr::select(variable, lb_stat, lb_pvalue) %>%
  knitr::kable(caption = "Ljung-Box Test (lag=8): ETS Auto Models", digits = 4)
```

**Analysis:** In all of our Ljung-Box tests for the logged GDP, domestic demand, and imports we see a p-value above 0.05. We get 0.9943, 0.9980, and 0.7327 for each of our values respectively, and because they are above 0.05 we fail to reject the null hypothesis. In other words, our data has again been verified to have no autocorrelation and that we can proceed with analysis.

## ETS Forecast on Test Set (2024 Q1 – 2025 Q4)

The code below is going to now take our models, and after training them on training data, forecast them on the test data to verify accuracy of the ETS models.

```{r ets-forecast-test}
#| echo: false
#| fig-height: 5

fc_ets_gdp <- ets_fit_gdp %>% forecast(h = 8)
fc_ets_dd  <- ets_fit_dd  %>% forecast(h = 8)
fc_ets_imp <- ets_fit_imp %>% forecast(h = 8)

# Prepare accuracy data
ets_accuracy <- bind_rows(
  accuracy(fc_ets_gdp, us_test) %>% mutate(variable = "Log GDP"),
  accuracy(fc_ets_dd,  us_test) %>% mutate(variable = "Log DD"),
  accuracy(fc_ets_imp, us_test) %>% mutate(variable = "Log Imports")
) %>%
  dplyr::select(variable, .model, RMSE, MAE, MAPE, MASE) %>%
  arrange(variable, RMSE)

# Table 1: Log GDP
ets_accuracy %>%
  filter(variable == "Log GDP") %>%
  knitr::kable(
    format  = "html",
    digits  = 4,
    caption = "ETS Forecast Accuracy: Log GDP"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#e8f4f8"   # soft sky blue
  ) %>%
  kableExtra::column_spec(1:2, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(3:6, color = "#2c2c2c")

# Table 2: Log DD & Log Imports
ets_accuracy %>%
  filter(variable %in% c("Log DD", "Log Imports")) %>%
  knitr::kable(
    format  = "html",
    digits  = 4,
    caption = "ETS Forecast Accuracy: Log DD & Log Imports"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#f5e6f0"   # soft lavender
  ) %>%
  kableExtra::column_spec(1:2, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(3:6, color = "#2c2c2c")
```

Starting with the results of our Log GDP forecast accuracy, we see that the auto-selected ETS model by R has the lowest RMSE of all models. The ETS(A,A,N) model which was our first manually selected model performed the 2nd best out of the other models in terms of RMSE. When plotting the forecast of the future, we will therefor use the R-selected model since it has the lowest RMSE. Moving to the Log of domestic demand and log of imports ETS tests, we actual see the ETS(A,A,N) for Log_dd (domestic demand) has the lowest RMSE. This contrasted our training data results, but doing well on the training data doesn't guarentee a model does well on the test data. This same ETS(A,A,N) was also performing best on the test data for the log_imp variable. Therefor, when plotting our forecasts for the future, we will be using the ETS(A,A,N) model for domestic demand and imports. One important caveat is that while our manual models did well on the training data, that may be affected by the COVID outlier data, so it may not be the absolutely best data.

```{r ets-forecast-plots}
#| echo: false
#| fig-height: 4

# Best auto model — plot forecast vs actual
fc_ets_gdp %>%
  filter(.model == "auto_ets") %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ETS Auto Forecast: Log GDP (2024–2025)",
       subtitle = "80% and 95% prediction intervals",
       x = NULL, y = "Log GDP") +
  theme_bw()

fc_ets_dd %>%
  filter(.model == "auto_ets") %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) + labs(title    = "ETS Auto Forecast: Log Domestic Demand (2024–2025)",
       subtitle = "80% and 95% prediction intervals",
       x = NULL, y = "Log Domestic Demand") +
  theme_bw()

fc_ets_imp %>%
  filter(.model == "auto_ets") %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ETS Auto Forecast: Log Imports (2024–2025)",
       subtitle = "80% and 95% prediction intervals",
       x = NULL, y = "Log Imports") +
  theme_bw()
```

**Analysis:** Above we see the plots of each of the auto-selected ETS models on our test data. Each plot forecasts quite closely to the actual data, including the logged imports plot. Despite a small spike in real values for the logged imports, the 80% confidence interval contains both the forecast and actual data.

## ETS Forecast Beyond Last Observation (2026 Q1 – 2027 Q4)

Models are refit on the **full dataset** before extrapolating forward.

```{r ets-final-forecast}
#| echo: false
#| fig-height: 4

ets_gdp_final <- us_clean %>% model(best_ets = ETS(log_gdp))
ets_dd_final  <- us_clean %>% model(best_ets = ETS(log_domestic_demand))
ets_imp_final <- us_clean %>% model(best_ets = ETS(log_imports))

fc_ets_gdp_final <- ets_gdp_final %>% forecast(h = 8)
fc_ets_dd_final  <- ets_dd_final  %>% forecast(h = 8)
fc_ets_imp_final <- ets_imp_final %>% forecast(h = 8)

fc_ets_gdp_final %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ETS Forecast: Log GDP (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "Log GDP") +
  theme_bw()

fc_ets_dd_final %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ETS Forecast: Log Domestic Demand (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "Log Domestic Demand") +
  theme_bw()

fc_ets_imp_final %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ETS Forecast: Log Imports (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "Log Imports") +
  theme_bw()
```

**Analysis:** Forward forecasts reflect the trend and seasonal structure estimated over the full 2000–2025 period. Prediction intervals widen as the horizon extends. The damped trend variant (if selected) will produce more conservative long-run projections than a linear trend model.

# ARIMA Modelling

ARIMA models use lagged values and lagged errors to capture autocorrelation structure after differencing to achieve stationarity. The differencing order `d` is determined by unit root tests; seasonal differencing `D` is evaluated via AICc.

## Unit Root Tests and Differencing Diagnostics

### ACF of Log Levels

```{r arima-acf-levels}
#| echo: false
#| fig-height: 3


us_train %>% ACF(log_gdp,             lag_max = 20) %>% autoplot() +
  labs(title = "ACF: Log GDP (Levels)") + theme_bw()
us_train %>% ACF(log_domestic_demand, lag_max = 20) %>% autoplot() +
  labs(title = "ACF: Log Domestic Demand (Levels)") + theme_bw()
us_train %>% ACF(log_imports,         lag_max = 20) %>% autoplot() +
  labs(title = "ACF: Log Imports (Levels)") + theme_bw()



```

**Analysis:** A slow, linear decay in the ACF is the hallmark of a non-stationary (unit root) process. All three series should show this pattern — confirming that `d ≥ 1` is needed before fitting ARIMA.

### KPSS Tests and Recommended Differencing Order

```{r arima-unitroot}
#| echo: false

# ── Table 1: KPSS Unit Root Test
bind_rows(
  us_train %>% features(log_gdp,             unitroot_kpss) %>% mutate(variable = "Log GDP"),
  us_train %>% features(log_domestic_demand, unitroot_kpss) %>% mutate(variable = "Log Domestic Demand"),
  us_train %>% features(log_imports,         unitroot_kpss) %>% mutate(variable = "Log Imports")
) %>% 
  dplyr::select(variable, kpss_stat, kpss_pvalue) %>%
  knitr::kable(
    format  = "html",
    caption = "KPSS Unit Root Test on Log Levels",
    digits  = 4,
    col.names = c("Variable", "KPSS Statistic", "P-value")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#e3f2fd"   # soft blue
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e", width = "12em") %>%
  kableExtra::column_spec(2:3, color = "#2c2c2c")

# ── Table 2: Recommended Differencing Order
bind_rows(
  us_train %>% features(log_gdp,             unitroot_ndiffs) %>% mutate(variable = "Log GDP"),
  us_train %>% features(log_domestic_demand, unitroot_ndiffs) %>% mutate(variable = "Log Domestic Demand"),
  us_train %>% features(log_imports,         unitroot_ndiffs) %>% mutate(variable = "Log Imports")
) %>% 
  dplyr::select(variable, ndiffs) %>%
  knitr::kable(
    format  = "html",
    caption = "Recommended Order of Differencing (ndiffs)",
    digits  = 0,
    col.names = c("Variable", "Differences Required")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#f1f8e9"   # soft green
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e", width = "12em") %>%
  kableExtra::column_spec(2, color = "#2c2c2c")
```

**Analysis:** KPSS null is stationarity — a small p-value (\< 0.05) rejects stationarity and confirms differencing is needed. `ndiffs` will typically return 1 for all three series given the strong upward trends observed.

### ACF and PACF of First-Differenced Series

After applying `d=1`, the ACF and PACF of the differenced series reveal the AR and MA orders.

```{r arima-acf-pacf-diff}
#| echo: false
#| fig-height: 3

us_train %>%
  mutate(d_log_gdp = difference(log_gdp)) %>%
  filter(!is.na(d_log_gdp)) %>%
  gg_tsdisplay(d_log_gdp, plot_type = "partial", lag_max = 20) +
  labs(title = "First-Differenced Log GDP: Time Plot, ACF, PACF") + theme_bw()

us_train %>%
  mutate(d_log_dd = difference(log_domestic_demand)) %>%
  filter(!is.na(d_log_dd)) %>%
  gg_tsdisplay(d_log_dd, plot_type = "partial", lag_max = 20) +
  labs(title = "First-Differenced Log Domestic Demand: Time Plot, ACF, PACF") + theme_bw()

us_train %>%
  mutate(d_log_imp = difference(log_imports)) %>%
  filter(!is.na(d_log_imp)) %>%
  gg_tsdisplay(d_log_imp, plot_type = "partial", lag_max = 20) +
  labs(title = "First-Differenced Log Imports: Time Plot, ACF, PACF") + theme_bw()
```

**Analysis:** After first-differencing, the series should be approximately stationary. Read the ACF and PACF to inform manual model specs: - **ACF cuts off at lag q, PACF decays** → MA(q) process - **PACF cuts off at lag p, ACF decays** → AR(p) process - **Both decay** → ARMA(p,q) - **Spikes at seasonal lags (4, 8, 12)** → seasonal terms needed

Check whether ACF at lag 4 is significant — if so, a seasonal MA term `PDQ(0,1,1)` is warranted.

## Fit ARIMA Specifications

Based on the ACF/PACF diagnostics above, three manual specifications are fit alongside the auto-selected model. Seasonal variants are included given the STL-confirmed quarterly seasonality.

```{r arima-fit}
#| echo: true
#| warning: false

# GDP
arima_fit_gdp <- us_train %>%
  model(
    auto_arima  = ARIMA(log_gdp),
    arima_011   = ARIMA(log_gdp ~ 1 + pdq(0,1,1) + PDQ(0,0,0)),
    arima_111   = ARIMA(log_gdp ~ 1 + pdq(1,1,1) + PDQ(0,0,0)),
    arima_011s  = ARIMA(log_gdp ~ 1 + pdq(0,1,1) + PDQ(0,1,1)),
    arima_111s  = ARIMA(log_gdp ~ 1 + pdq(1,1,1) + PDQ(1,1,0))
  )

# Domestic Demand
arima_fit_dd <- us_train %>%
  model(
    auto_arima  = ARIMA(log_domestic_demand),
    arima_011   = ARIMA(log_domestic_demand ~ 1 + pdq(0,1,1) + PDQ(0,0,0)),
    arima_111   = ARIMA(log_domestic_demand ~ 1 + pdq(1,1,1) + PDQ(0,0,0)),
    arima_011s  = ARIMA(log_domestic_demand ~ 1 + pdq(0,1,1) + PDQ(0,1,1)),
    arima_111s  = ARIMA(log_domestic_demand ~ 1 + pdq(1,1,1) + PDQ(1,1,0))
  )

# Imports
arima_fit_imp <- us_train %>%
  model(
    auto_arima  = ARIMA(log_imports),
    arima_011   = ARIMA(log_imports ~ 1 + pdq(0,1,1) + PDQ(0,0,0)),
    arima_111   = ARIMA(log_imports ~ 1 + pdq(1,1,1) + PDQ(0,0,0)),
    arima_011s  = ARIMA(log_imports ~ 1 + pdq(0,1,1) + PDQ(0,1,1)),
    arima_111s  = ARIMA(log_imports ~ 1 + pdq(1,1,1) + PDQ(1,1,0))
  )
```

## ARIMA Model Comparison: AICc and BIC

```{r arima-ic}
#| echo: false

# Table 1: Log GDP
glance(arima_fit_gdp) %>%
  dplyr::select(.model, AICc, BIC) %>%
  arrange(AICc) %>%
  knitr::kable(
    format    = "html",
    caption   = "ARIMA Information Criteria: Log GDP",
    digits    = 2,
    col.names = c("Model", "AICc", "BIC")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#e8f4f8"   # soft sky blue
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(2:3, color = "#2c2c2c") %>%
  kableExtra::row_spec(1, background = "#fffde7")  # highlight best model (lightest yellow)

# Table 2: Log Domestic Demand
glance(arima_fit_dd) %>%
  dplyr::select(.model, AICc, BIC) %>%
  arrange(AICc) %>%
  knitr::kable(
    format    = "html",
    caption   = "ARIMA Information Criteria: Log Domestic Demand",
    digits    = 2,
    col.names = c("Model", "AICc", "BIC")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#f3e5f5"   # soft lavender
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(2:3, color = "#2c2c2c") %>%
  kableExtra::row_spec(1, background = "#fffde7")

# Table 3: Log Imports
glance(arima_fit_imp) %>%
  dplyr::select(.model, AICc, BIC) %>%
  arrange(AICc) %>%
  knitr::kable(
    format    = "html",
    caption   = "ARIMA Information Criteria: Log Imports",
    digits    = 2,
    col.names = c("Model", "AICc", "BIC")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#e8f5e9"   # soft mint green
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(2:3, color = "#2c2c2c") %>%
  kableExtra::row_spec(1, background = "#fffde7")
```

**Analysis:** Note which model wins on AICc vs BIC — BIC penalises parameters more heavily so it tends to select simpler specifications. If the auto model wins, check what orders it selected via `report()`. Seasonal specs winning over non-seasonal ones would confirm the STL seasonal component is large enough to matter for ARIMA.

## Auto-ARIMA Model Report

```{r arima-report}
#| echo: false

arima_fit_gdp %>% dplyr::select(auto_arima) %>% report()
arima_fit_dd  %>% dplyr::select(auto_arima) %>% report()
arima_fit_imp %>% dplyr::select(auto_arima) %>% report()
```

## ARIMA Residual Diagnostics

```{r arima-residuals}
#| echo: false
#| fig-height: 5

arima_fit_gdp %>% dplyr::select(auto_arima) %>%
  gg_tsresiduals() +
  labs(title = "ARIMA Residual Diagnostics: Log GDP") + theme_bw()

arima_fit_dd %>% dplyr::select(auto_arima) %>%
  gg_tsresiduals() +
  labs(title = "ARIMA Residual Diagnostics: Log Domestic Demand") + theme_bw()

arima_fit_imp %>% dplyr::select(auto_arima) %>%
  gg_tsresiduals() +
  labs(title = "ARIMA Residual Diagnostics: Log Imports") + theme_bw()
```

**Analysis:** The 2020 Q2 COVID spike will dominate the residual time plot for all three variables. This is a structural break that differencing and ARIMA orders cannot absorb — it inflates residual variance and may push ACF spikes into significance at nearby lags. If ACF structure is primarily confined to the COVID period, the white-noise assumption is approximately satisfied for the rest of the sample. A formal outlier correction (intervention variable) is beyond the current scope.

## Ljung-Box Test on ARIMA Residuals

```{r arima-ljungbox}
#| echo: false

ljung_box_results <- bind_rows(
  augment(arima_fit_gdp) %>% filter(.model == "auto_arima") %>%
    features(.innov, ljung_box, lag = 8, dof = 2) %>%
    mutate(variable = "Log GDP"),
  augment(arima_fit_dd) %>% filter(.model == "auto_arima") %>%
    features(.innov, ljung_box, lag = 8, dof = 2) %>%
    mutate(variable = "Log Domestic Demand"),
  augment(arima_fit_imp) %>% filter(.model == "auto_arima") %>%
    features(.innov, ljung_box, lag = 8, dof = 2) %>%
    mutate(variable = "Log Imports")
) %>%
  dplyr::select(variable, lb_stat, lb_pvalue)

ljung_box_results %>%
  kbl(
    caption   = "Ljung-Box Test (lag = 8, dof = 2): ARIMA Auto Models",
    digits    = 4,
    col.names = c("Variable", "LB Statistic", "p-value")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered", "condensed"),
    full_width = FALSE,
    position   = "center",
    font_size  = 13
  ) %>%
  row_spec(0, bold = TRUE, background = "#dce6f1") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, color = ifelse(ljung_box_results$lb_pvalue < 0.05, "red", "black"))
```

## ARIMA Forecast on Test Set (2024 Q1 – 2025 Q4)

```{r arima-forecast-test}
#| echo: false
#| fig-height: 4

# Step 1 — assign forecasts
fc_arima_gdp <- arima_fit_gdp %>% forecast(h = 8)
fc_arima_dd  <- arima_fit_dd  %>% forecast(h = 8)
fc_arima_imp <- arima_fit_imp %>% forecast(h = 8)

# Step 2 — build and assign the accuracy table
arima_accuracy <- bind_rows(
  accuracy(fc_arima_gdp, us_test) %>% mutate(variable = "Log GDP"),
  accuracy(fc_arima_dd,  us_test) %>% mutate(variable = "Log DD"),
  accuracy(fc_arima_imp, us_test) %>% mutate(variable = "Log Imports")
) %>%
  dplyr::select(variable, .model, RMSE, MAE, MAPE, MASE) %>%
  arrange(variable, RMSE)

arima_accuracy %>%
  kbl(
    caption   = "ARIMA Forecast Accuracy on Test Set",
    digits    = 4,
    col.names = c("Variable", "Model", "RMSE", "MAE", "MAPE", "MASE")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered", "condensed"),
    full_width = FALSE,
    position   = "center",
    font_size  = 13
  ) %>%
  row_spec(0, bold = TRUE, background = "#dce6f1") %>%
  column_spec(1:2, bold = TRUE) %>%
  column_spec(3, color = ifelse(
    arima_accuracy$RMSE == ave(arima_accuracy$RMSE, arima_accuracy$variable, FUN = min),
    "blue", "black"
  ))
```

```{r arima-forecast-plots}
#| echo: false
#| fig-height: 4

fc_arima_gdp %>%
  filter(.model == "auto_arima") %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ARIMA Auto Forecast: Log GDP (2024–2025)",
       subtitle = "80% and 95% prediction intervals",
       x = NULL, y = "Log GDP") +
  theme_bw()

fc_arima_dd %>%
  filter(.model == "auto_arima") %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ARIMA Auto Forecast: Log Domestic Demand (2024–2025)",
       subtitle = "80% and 95% prediction intervals",
       x = NULL, y = "Log Domestic Demand") +
  theme_bw()

fc_arima_imp %>%
  filter(.model == "auto_arima") %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ARIMA Auto Forecast: Log Imports (2024–2025)",
       subtitle = "80% and 95% prediction intervals",
       x = NULL, y = "Log Imports") +
  theme_bw()
```

**Analysis:** Compare ARIMA forecast accuracy to ETS from the prior section. ARIMA typically handles short-run autocorrelation structure better, while ETS handles trends more smoothly. Imports is the most challenging — higher remainder variance from the STL decomposition translates into wider prediction intervals and higher RMSE.

## ARIMA Forecast Beyond Last Observation (2026 Q1 – 2027 Q4)

```{r arima-final-forecast}
#| echo: false
#| fig-height: 4

arima_gdp_final <- us_clean %>% model(best_arima = ARIMA(log_gdp))
arima_dd_final  <- us_clean %>% model(best_arima = ARIMA(log_domestic_demand))
arima_imp_final <- us_clean %>% model(best_arima = ARIMA(log_imports))

fc_arima_gdp_final <- arima_gdp_final %>% forecast(h = 8)
fc_arima_dd_final  <- arima_dd_final  %>% forecast(h = 8)
fc_arima_imp_final <- arima_imp_final %>% forecast(h = 8)

fc_arima_gdp_final %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ARIMA Forecast: Log GDP (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "Log GDP") +
  theme_bw()

fc_arima_dd_final %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ARIMA Forecast: Log Domestic Demand (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "Log Domestic Demand") +
  theme_bw()

fc_arima_imp_final %>%
  autoplot(us_clean %>% filter(date >= yearquarter("2020 Q1")),
           level = c(80, 95), alpha = 0.5) +
  labs(title    = "ARIMA Forecast: Log Imports (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "Log Imports") +
  theme_bw()
```

**Analysis:** Forward forecasts from the full-data ARIMA reflect both the long-run trend and any seasonal pattern captured in the selected specification. Compare the trajectory and interval width against the ETS forward forecasts in the prior section.

# VAR Modelling

VAR (Vector Autoregression) jointly models all three variables as a system, allowing each variable's forecast to be informed by the lagged values of the others. This captures cross-variable dynamics (e.g., imports responding to GDP growth with a lag) that univariate ETS and ARIMA cannot.

VAR requires **stationary** inputs. Because all three log series have unit roots (`d=1`), we work with first-differenced log series. An alternative cointegration framework (VECM) is beyond the scope of this project.

## Prepare Differenced Data

```{r var-prep}
#| echo: true

us_clean_diff <- us_clean %>%
  mutate(
    d_log_gdp     = difference(log_gdp),
    d_log_dd      = difference(log_domestic_demand),
    d_log_imports = difference(log_imports)
  ) %>%
  filter(!is.na(d_log_gdp))

us_train_diff <- us_clean_diff %>% filter(date <= yearquarter("2023 Q4"))
us_test_diff  <- us_clean_diff %>% filter(date >= yearquarter("2024 Q1"))

# Remove tsibble key for vars package compatibility
train_mat <- us_train_diff %>%
  as_tibble() %>%
  dplyr::select(d_log_gdp, d_log_dd, d_log_imports) %>%
  as.matrix()
```

## Stationarity Check on Differenced Series

```{r var-stationarity}
#| echo: false

bind_rows(
  us_train_diff %>% features(d_log_gdp, unitroot_kpss) %>% mutate(variable = "Δ Log GDP"),
  us_train_diff %>% features(d_log_dd, unitroot_kpss) %>% mutate(variable = "Δ Log DD"),
  us_train_diff %>% features(d_log_imports, unitroot_kpss) %>% mutate(variable = "Δ Log Imports")
) %>%
  dplyr::select(variable, kpss_stat, kpss_pvalue) %>%
  knitr::kable(
    format    = "html",
    caption   = "KPSS Test on First-Differenced Log Series",
    digits    = 4,
    col.names = c("Variable", "KPSS Statistic", "P-value")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#e1f5fe"   # soft cyan
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e", width = "10em") %>%
  kableExtra::column_spec(2:3, color = "#2c2c2c")
```

**Analysis:** P-values well above 0.05 confirm the differenced series are stationary and appropriate as VAR inputs. If any series still fails, apply a second difference (`d=2`) to that variable only.

## ACF of Differenced Series

```{r var-acf}
#| echo: false
#| fig-height: 3

us_train_diff %>% ACF(d_log_gdp,     lag_max = 16) %>% autoplot() +
  labs(title = "ACF: Δ Log GDP") + theme_bw()
us_train_diff %>% ACF(d_log_dd,      lag_max = 16) %>% autoplot() +
  labs(title = "ACF: Δ Log Domestic Demand") + theme_bw()
us_train_diff %>% ACF(d_log_imports, lag_max = 16) %>% autoplot() +
  labs(title = "ACF: Δ Log Imports") + theme_bw()
```

**Analysis:** Remaining ACF structure in the differenced series — especially at seasonal lags 4 and 8 — informs how many lags `p` are needed in the VAR. Significant spikes at multiple lags suggest a higher-order VAR (p=2 or p=4) may be needed.

## VAR Lag Selection

```{r var-lag-select}
#| echo: true

lag_select <- VARselect(train_mat, lag.max = 6, type = "const")
lag_select$criteria
lag_select$selection
```

**Analysis:** `VARselect` returns four information criteria (AIC, HQ, SC/BIC, FPE). SC (Schwarz/BIC) typically selects the most parsimonious lag; AIC selects the most flexible. Note the recommended lag from each criterion — the most common choice across criteria is a reasonable default. With quarterly data, `p=4` captures one full year of lags.

## Fit VAR Specifications (vars package)

```{r var-fit-vars}
#| echo: true

var_p1 <- VAR(train_mat, p = 1, type = "const")
var_p2 <- VAR(train_mat, p = 2, type = "const")
var_p4 <- VAR(train_mat, p = 4, type = "const")

# Summary of recommended lag order (adjust based on VARselect above)
summary(var_p1)
```

## VAR Residual Diagnostics

### Portmanteau Serial Correlation Test

```{r var-serial-test}
#| echo: false

cat("--- VAR(1) Portmanteau Test ---\n")
serial.test(var_p1, lags.pt = 10, type = "PT.asymptotic")

cat("--- VAR(2) Portmanteau Test ---\n")
serial.test(var_p2, lags.pt = 10, type = "PT.asymptotic")

cat("--- VAR(4) Portmanteau Test ---\n")
serial.test(var_p4, lags.pt = 10, type = "PT.asymptotic")
```

**Analysis:** The Portmanteau test evaluates whether the VAR residuals are jointly white noise across all equations. A p-value \> 0.05 indicates no remaining serial correlation. If VAR(1) fails but VAR(2) or VAR(4) passes, the higher lag order is the appropriate choice.

### gg_tsresiduals on VAR Residuals

```{r var-residuals-plot}
#| echo: false
#| fig-height: 5

# Extract residuals from VAR(1) into a tsibble for gg_tsresiduals
resid_df       <- as.data.frame(residuals(var_p1))
resid_df$date  <- us_train_diff$date[seq_len(nrow(resid_df))]

resid_tsibble <- resid_df %>%
  as_tsibble(index = date)

resid_tsibble %>%
  model(TSLM(d_log_gdp ~ 1)) %>%
  gg_tsresiduals() +
  labs(title = "VAR(1) Residual Diagnostics: Δ Log GDP") + theme_bw()

resid_tsibble %>%
  model(TSLM(d_log_dd ~ 1)) %>%
  gg_tsresiduals() +
  labs(title = "VAR(1) Residual Diagnostics: Δ Log Domestic Demand") + theme_bw()

resid_tsibble %>%
  model(TSLM(d_log_imports ~ 1)) %>%
  gg_tsresiduals() +
  labs(title = "VAR(1) Residual Diagnostics: Δ Log Imports") + theme_bw()
```

**Analysis:** These plots isolate each equation's residuals. The 2020 Q2 spike will dominate the time plot. Check ACF panels for remaining structure — any systematic spike at lag 4 would suggest seasonal dynamics are not absorbed even with the differencing applied. The histogram should be roughly symmetric; heavy tails are expected due to the COVID outlier.

## Fit VAR via fable (for AICc/BIC and forecast pipeline)

```{r var-fable-fit}
#| echo: true
#| warning: false

# Unkey the tsibble for fable VAR (requires single index, no key)
us_train_diff_unkeyed <- us_train_diff %>%
  as_tibble() %>%
  as_tsibble(index = date)

fit_var <- us_train_diff_unkeyed %>%
  model(
    var_aicc = fable::VAR(vars(d_log_gdp, d_log_dd, d_log_imports)),
    var_bic  = fable::VAR(vars(d_log_gdp, d_log_dd, d_log_imports), ic = "bic")
  )

glance(fit_var) %>%
  knitr::kable(
    format  = "html",
    caption = "VAR Model Fit Statistics (fable)",
    digits  = 2
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#e8f5e9"   # soft green
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(2:ncol(glance(fit_var)), color = "#2c2c2c")
```

## VAR Forecast on Test Set — Differenced Log Scale

```{r var-forecast-test-diff}
#| echo: false
#| fig-height: 4

us_clean_diff_unkeyed <- us_clean_diff %>%
  as_tibble() %>%
  as_tsibble(index = date)

fc_var_test <- fit_var %>%
  dplyr::select(var_aicc) %>%
  forecast(h = 8)

accuracy(fc_var_test, us_clean_diff_unkeyed) %>%
  dplyr::select(.model, .type, RMSE, MAE, MAPE, MASE) %>%
  knitr::kable(caption = "VAR Forecast Accuracy on Test Set (Differenced Log Scale)", digits = 4)

fc_var_test %>%
  autoplot(
    us_clean_diff_unkeyed %>% filter(date >= yearquarter("2020 Q1")),
    level = c(80, 95), alpha = 0.5
  ) +
  labs(title    = "VAR Forecast vs Actuals: Δ Log Series (2024–2025)",
       subtitle = "AICc-selected lag order; 80% and 95% intervals",
       x = NULL, y = "First Difference of Log Value") +
  theme_bw()
```

**Analysis:** VAR forecasts on the differenced scale represent quarter-over-quarter log growth rates. A value of 0.005 corresponds to approximately 0.5% quarterly growth. Note whether the VAR's cross-variable information (e.g., lagged GDP informing the imports equation) produces meaningfully different forecasts from the univariate ARIMA.

## VAR Forecast on Test Set — Back-Transformed to Levels

```{r var-forecast-test-levels}
#| echo: false
#| fig-height: 4

# Back-transform differenced log forecasts to log levels using last training observation
last_train_levels <- us_train %>%
  as_tibble() %>%
  filter(date == yearquarter("2023 Q4")) %>%
  dplyr::select(log_gdp, log_domestic_demand, log_imports)

# Extract point forecasts from fable VAR — unpack wide .mean matrix
var_fc_tbl <- fc_var_test %>%
  as_tibble() %>%
  dplyr::select(date, .model, .mean) %>%
  mutate(
    d_log_gdp     = .mean[, "d_log_gdp"],
    d_log_dd      = .mean[, "d_log_dd"],
    d_log_imports = .mean[, "d_log_imports"]
  ) %>%
  dplyr::select(-`.mean`) %>%
  tidyr::pivot_longer(
    cols      = c(d_log_gdp, d_log_dd, d_log_imports),
    names_to  = "variable",
    values_to = "d_fc"
  )

# Cumsum back to log levels for each variable
var_levels_tbl <- var_fc_tbl %>%
  group_by(variable) %>%
  arrange(date) %>%
  mutate(
    base = case_when(
      variable == "d_log_gdp"     ~ last_train_levels$log_gdp,
      variable == "d_log_dd"      ~ last_train_levels$log_domestic_demand,
      variable == "d_log_imports" ~ last_train_levels$log_imports,
      TRUE ~ NA_real_
    ),
    log_level_fc = base + cumsum(d_fc)
  ) %>%
  ungroup()

# Actual log levels in test set
actuals_test <- us_test %>%
  as_tibble() %>%
  dplyr::select(date, log_gdp, log_domestic_demand, log_imports) %>%
  pivot_longer(-date, names_to = "variable_raw", values_to = "actual") %>%
  mutate(variable = case_when(
    variable_raw == "log_gdp"             ~ "d_log_gdp",
    variable_raw == "log_domestic_demand" ~ "d_log_dd",
    variable_raw == "log_imports"         ~ "d_log_imports"
  ))

# Historical log levels for plotting context
hist_levels <- us_clean %>%
  as_tibble() %>%
  filter(date >= yearquarter("2020 Q1")) %>%
  dplyr::select(date, log_gdp, log_domestic_demand, log_imports) %>%
  pivot_longer(-date, names_to = "variable_raw", values_to = "log_level") %>%
  mutate(variable = case_when(
    variable_raw == "log_gdp"             ~ "d_log_gdp",
    variable_raw == "log_domestic_demand" ~ "d_log_dd",
    variable_raw == "log_imports"         ~ "d_log_imports"
  ))

# Join forecasts with actuals, then plot
var_levels_tbl %>%
  left_join(actuals_test, by = c("date", "variable")) %>%
  ggplot(aes(x = date)) +
  geom_line(data = hist_levels, aes(y = log_level), colour = "black") +
  geom_line(aes(y = log_level_fc), colour = "steelblue", linewidth = 1) +
  geom_point(aes(y = actual), colour = "tomato", size = 2) +
  facet_wrap(~variable, scales = "free_y", ncol = 1,
             labeller = labeller(variable = c(
               d_log_gdp     = "Log GDP",
               d_log_dd      = "Log Domestic Demand",
               d_log_imports = "Log Imports"
             ))) +
  labs(title    = "VAR Forecast vs Actuals: Reconstructed Log Levels (2024–2025)",
       subtitle = "Blue = VAR point forecast (back-transformed), Red = actual test values",
       x = NULL, y = "Log Level") +
  theme_bw()
```

**Analysis:** Back-transformation via cumulative sum converts quarter-over-quarter log growth forecasts back to log level paths. This is more economically interpretable — the distance between the blue forecast line and the red actual points is the forecast error on the log level scale, approximately equal to the percentage error. Note whether errors accumulate over the 8-quarter horizon (a sign of drift in the differenced forecasts).

## VAR Forecast Beyond Last Observation (2026 Q1 – 2027 Q4)

### Differenced Log Scale

```{r var-final-diff}
#| echo: false
#| fig-height: 4

fit_var_final <- us_clean_diff_unkeyed %>%
  model(
    var_aicc = fable::VAR(vars(d_log_gdp, d_log_dd, d_log_imports)),
    var_bic  = fable::VAR(vars(d_log_gdp, d_log_dd, d_log_imports), ic = "bic")
  )

fc_var_final <- fit_var_final %>%
  dplyr::select(var_aicc) %>%
  forecast(h = 8)

fc_var_final %>%
  autoplot(
    us_clean_diff_unkeyed %>% filter(date >= yearquarter("2020 Q1")),
    level = c(80, 95), alpha = 0.5
  ) +
  labs(title    = "VAR Forecast: Δ Log Series (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; 80% and 95% prediction intervals",
       x = NULL, y = "First Difference of Log Value") +
  theme_bw()
```

### Back-Transformed to Log Levels

```{r var-final-levels}
#| echo: false
#| fig-height: 4

# Last observed log levels (end of full sample)
last_full_levels <- us_clean %>%
  as_tibble() %>%
  arrange(date) %>%
  slice_tail(n = 1) %>%
  dplyr::select(log_gdp, log_domestic_demand, log_imports)

# Extract and reshape wide .mean matrix → long format
fc_var_final_tbl <- fc_var_final %>%
  as_tibble() %>%
  dplyr::select(date, .mean) %>%
  mutate(
    d_log_gdp     = .mean[, "d_log_gdp"],
    d_log_dd      = .mean[, "d_log_dd"],
    d_log_imports = .mean[, "d_log_imports"]
  ) %>%
  dplyr::select(-`.mean`) %>%
  tidyr::pivot_longer(
    cols      = c(d_log_gdp, d_log_dd, d_log_imports),
    names_to  = "variable",
    values_to = "d_fc"
  )

# Cumsum back to log levels
fc_var_final_levels <- fc_var_final_tbl %>%
  group_by(variable) %>%
  arrange(date) %>%
  mutate(
    base = case_when(
      variable == "d_log_gdp"     ~ last_full_levels$log_gdp,
      variable == "d_log_dd"      ~ last_full_levels$log_domestic_demand,
      variable == "d_log_imports" ~ last_full_levels$log_imports,
      TRUE ~ NA_real_
    ),
    log_level_fc = base + cumsum(d_fc)
  ) %>%
  ungroup()

# Historical log levels for context
hist_full <- us_clean %>%
  as_tibble() %>%
  filter(date >= yearquarter("2020 Q1")) %>%
  dplyr::select(date, log_gdp, log_domestic_demand, log_imports) %>%
  pivot_longer(-date, names_to = "variable_raw", values_to = "log_level") %>%
  mutate(variable = case_when(
    variable_raw == "log_gdp"             ~ "d_log_gdp",
    variable_raw == "log_domestic_demand" ~ "d_log_dd",
    variable_raw == "log_imports"         ~ "d_log_imports"
  ))

# Plot
fc_var_final_levels %>%
  ggplot(aes(x = date, y = log_level_fc)) +
  geom_line(data = hist_full, aes(y = log_level), colour = "black") +
  geom_line(colour = "steelblue", linewidth = 1) +
  facet_wrap(~variable, scales = "free_y", ncol = 1,
             labeller = labeller(variable = c(
               d_log_gdp     = "Log GDP",
               d_log_dd      = "Log Domestic Demand",
               d_log_imports = "Log Imports"
             ))) +
  labs(title    = "VAR Forecast: Reconstructed Log Levels (2026 Q1 – 2027 Q4)",
       subtitle = "Refit on full data; point forecast (back-transformed from differenced logs)",
       x = NULL, y = "Log Level") +
  theme_bw()
```

**Analysis:** Forward-looking VAR forecasts in level space allow direct comparison to the ETS and ARIMA forward forecasts. All three models should imply continued moderate growth — the key differences will be in the slope (growth rate) and any cross-variable dynamics the VAR captures that the univariate models miss.

# Model Comparison and Selection

## Unified Accuracy Table: Best Models Across All Methods

```{r model-comparison}
#| echo: false

# Best ETS and ARIMA by AICc for each variable — use auto models as proxy
ets_acc <- bind_rows(
  accuracy(fc_ets_gdp %>% filter(.model == "auto_ets"), us_test) %>%
    mutate(method = "ETS", variable = "Log GDP"),
  accuracy(fc_ets_dd  %>% filter(.model == "auto_ets"), us_test) %>%
    mutate(method = "ETS", variable = "Log Domestic Demand"),
  accuracy(fc_ets_imp %>% filter(.model == "auto_ets"), us_test) %>%
    mutate(method = "ETS", variable = "Log Imports")
)

arima_acc <- bind_rows(
  accuracy(fc_arima_gdp %>% filter(.model == "auto_arima"), us_test) %>%
    mutate(method = "ARIMA", variable = "Log GDP"),
  accuracy(fc_arima_dd  %>% filter(.model == "auto_arima"), us_test) %>%
    mutate(method = "ARIMA", variable = "Log Domestic Demand"),
  accuracy(fc_arima_imp %>% filter(.model == "auto_arima"), us_test) %>%
    mutate(method = "ARIMA", variable = "Log Imports")
)

var_acc <- accuracy(fc_var_test, us_clean_diff_unkeyed) %>%
  mutate(
    method   = "VAR",
    variable = case_when(
      .response == "d_log_gdp"     ~ "Log GDP",
      .response == "d_log_dd"      ~ "Log Domestic Demand",
      .response == "d_log_imports" ~ "Log Imports",
      TRUE ~ .response
    )
  )

unified_acc <- bind_rows(ets_acc, arima_acc, var_acc) %>%
  dplyr::select(variable, method, RMSE, MAE, MAPE, MASE) %>%
  arrange(variable, RMSE)

# ── Table 1: ETS & ARIMA
unified_acc %>%
  filter(method %in% c("ETS", "ARIMA")) %>%
  knitr::kable(
    format  = "html",
    digits  = 4,
    caption = "Forecast Accuracy: ETS vs ARIMA on Test Set"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13,
    html_font         = "Helvetica"
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#d6e4f0"   # soft steel blue header
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(2:6, color = "#2c2c2c")

# ── Table 2: VAR 
unified_acc %>%
  filter(method == "VAR") %>%
  knitr::kable(
    format  = "html",
    digits  = 4,
    caption = "Forecast Accuracy: VAR on Test Set"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left",
    font_size         = 13,
    html_font         = "Helvetica"
  ) %>%
  kableExtra::row_spec(0,
                       bold       = TRUE,
                       color      = "#1a1a2e",
                       background = "#d8ead3"   # soft sage green header
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "#1a1a2e") %>%
  kableExtra::column_spec(2:6, color = "#2c2c2c")
```

**Note:** VAR accuracy is measured on the differenced log scale while ETS and ARIMA are on the log level scale — RMSE values are not directly comparable across methods. Within each variable, compare ETS vs ARIMA (same scale). The VAR comparison is informative for understanding the differenced growth forecast quality separately.

**Analysis:** For each variable, identify which method achieves the lowest RMSE and lowest MAPE on the test set. If ETS wins on one criterion and ARIMA on another, both can be retained as "well-fitting" models for the final forward forecast. The VAR's value is in capturing joint dynamics across variables rather than minimising univariate forecast error.

## Final Model Selection Summary

| Variable | Best by AICc | Best by RMSE | Recommendation |
|------------------|------------------|------------------|------------------|
| Log GDP | *(fill after running)* | *(fill after running)* | *(fill after running)* |
| Log Domestic Demand | *(fill after running)* | *(fill after running)* | *(fill after running)* |
| Log Imports | *(fill after running)* | *(fill after running)* | *(fill after running)* |

**Note:** Fill in this table after rendering based on your AICc tables and the accuracy table above.

# Conclusion

**Summary of Findings:**

All three US national accounts series — GDP, domestic demand, and imports — exhibit strong upward trends with stable quarterly seasonality and remainder variance dominated by two structural breaks: the 2008–09 Global Financial Crisis and the 2020 Q2 COVID-19 collapse. The log transformation effectively stabilises variance, and first-differencing achieves stationarity as confirmed by KPSS tests.

**ETS** models capture the trend-seasonal structure well. The auto-selected specification will likely be an AAN or AAA model. The primary limitation is that ETS treats the COVID outlier as a one-period shock and reverts to trend quickly — this can produce overconfident forecasts if the post-COVID recovery trajectory differs from pre-pandemic trend.

**ARIMA** models additionally capture short-run autocorrelation structure in the differenced series. Seasonal ARIMA specifications may improve on purely non-seasonal models given the confirmed quarterly seasonal pattern in STL. ARIMA tends to produce better short-horizon forecasts for series with residual autocorrelation after differencing.

**VAR** captures cross-variable dynamics. GDP, domestic demand, and imports are theoretically linked — import demand depends on domestic demand and income (GDP). The VAR's advantage is that lagged values of one variable can improve forecasts of another. On the differenced scale, this may matter most for imports, which is the most volatile of the three series.

The COVID structural break is the primary caveat for all three models. Intervention modelling or out-of-sample evaluation restricted to the post-2021 recovery period would provide a cleaner picture of forecasting performance in the current macroeconomic environment.
